---
title: "Homework 2"
output: github_document
date: "2026-10-06"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

### Problem 1

**Step 1: Import and Clean Dataset**
```{r read_csv}
##Read and clean the data (copied from assignment)

nytransit_df = read_csv(file = "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                                na = c("", "NA", ".")) #cleans missing data

names(nytransit_df)#overview of all variables

nytransit_df = janitor::clean_names(nytransit_df)#convert to snake case
names(nytransit_df)#checks if above line ran properly
```
Double checking to see if cleaning data done properly:
```{r checking_data2}
nytransit_df
```

**Step 2: Data Manipulation**
```{r select}
##retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character to a logical variable 

nytransit_df <- 
  nytransit_df |> 
  select(line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada) |>
mutate(
    entry = recode_values(
      tolower(entry),  # to lower handles mixed case like "NO" or "Yes"
      "yes" ~ TRUE,
      "no"  ~ FALSE,
    ) #recode_values used instead of case_match to fix warning that came up
  )
```
Double checking to see if data manipulation done properly:
```{r checking_data}
nytransit_df #entry successfully converted to logical variable, all intended columns kept

nrow(nytransit_df)
ncol(nytransit_df)
```
**About this Dataset:**
(Q: Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?)

This data set (nytransit_df) has the following variables:line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. We first read and tidied the data and then manipulated it so only the desired columns as mentioned were kept as well as converting the 'entry' variables from characters to logical variables. The dataset was cleaned in three steps, first is converting any missing data to 'NA', then using the janitor function to standardize column numbes into snakecase and finally checking the resulting tibble to see if rows need to be skipped to find there were none. The dimension as mentioned in the last checking step is 1868 by 19 (1868 rows or entries by 19 columns or variables). The data is tidy based on all the steps done above. 

**Step 3: info retrieval**
```{r retrieval}
#counting distinct stations in nytransit_df
distinct_stations <- nytransit_df |> 
  distinct(station_name, line) |> 
  nrow()

distinct_stations

ada_stations <- nytransit_df |> 
  filter(ada == TRUE) |>   
  distinct(station_name, line) |> 
  nrow()

ada_stations

nytransit_df |> 
  filter(vending == "NO") |> 
  summarize(mean(entry))
```
**Answer the following questions with step 3:**
How many distinct stations are there? 
There are 465 distinct stations.

How many stations are ADA compliant?
84 stations are ADA compliant.

What proportion of station entrances / exits without vending allow entrance?
About 0.3770 or 37.70% of station entrances/exits without vending allow entrance. 

**Step 4: data reformatting**
```{r reformat}
# 1. Reformat data: convert all route columns to character first, then pivot
nytransit_tidy <- nytransit_df |> 
  mutate(across(route1:route11, as.character)) |> 
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  )

# 2. How many distinct stations serve the A train?
a_train_stations <- nytransit_tidy |> 
  filter(route_name == "A") |> 
  distinct(station_name, line) |> 
  nrow()

a_train_stations

# 3. Of the stations that serve the A train, how many are ADA compliant?
a_train_ada_stations <- nytransit_tidy |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(station_name, line) |> 
  nrow()

print(a_train_ada_stations)

```
**Reformatted data interpretation:**
60 distinct stations serve the A train and of stations that serve the A train, 17 are ADA compliant. 


### Problem 2

**Step 1: Read and clean Mr. Trash Wheel Sheet:**
```{r import}
#Read and clean the Mr. Trash Wheel sheet:specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel, use reasonable variable names, omit rows that do not include dumpster-specific data, round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer) (copied from assignment)

mrtw_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                     sheet = "Mr. Trash Wheel",
                     range = "A2:N742",
                     na = c("", "NA", "."))|>
  janitor::clean_names() |>
  mutate(sports_balls = as.integer(round(sports_balls)),
         trash_wheel = "Mr. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles)
         )

mrtw_df #this is to check for non-data entries, we see a '...15' and '...16' columns which should be ommitted, we also see that the last row doesn't include dumpster-specific data (These 'issues' are now fixed, we also have a clean dataset)
```
**Step 2: Read and clean Prof. Trash Wheeland Gwynnda Sheet:**
```{r import2}
#Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda (copied from assignment)

proftw_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                     sheet = "Professor Trash Wheel",
                     range = "A2:M139",
                     na = c("", "NA", "."))|>
  janitor::clean_names() |>
  mutate(
    trash_wheel = "Professor Trash Wheel",
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles)
  )

proftw_df #this is to check cleanliness of df,prof tw has no sports_balls column as a note

gwynnda_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                     sheet = "Gwynnda the Good Wheel of the W",
                     range = "A2:L394",
                     na = c("", "NA", "."))|>
  janitor::clean_names() |>
  mutate(
    trash_wheel = "Gwynnda",
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles)
  )

gwynnda_df #this is to check cleanliness of df,gwynnda has no sports_balls column as a note
```

**Step 3: Combine the three datasets**
```{r combine}
#To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining. This new variable is trash_wheel. 
tw_tidy <- bind_rows(mrtw_df, proftw_df, gwynnda_df) |>
  relocate(trash_wheel)

#View combined dataset
tw_tidy #shows a tidied, combined dataset with the variable #trash_wheel showing which dumpster the data belonged to
```

**Step 4: Info retrieval**
```{r retrieval_p2}
# 1. Total weight of trash collected by Professor Trash Wheel (in tons)
prof_weight <- tw_tidy |> 
  filter(trash_wheel == "Professor Trash Wheel") |> 
  summarize(total_weight_tons = sum(weight_tons, na.rm = TRUE))

prof_weight

# 2. Total cigarette butts collected by Gwynnda in June 2022
gwynnda_cigs <- tw_tidy |> 
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"  
  ) |> 
  summarize(total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE))

gwynnda_cigs
```
## About the data: 
This new dataset is `r nrow(tw_tidy)`rows by `r ncol(tw_tidy)` columns and contains `r nrow(tw_tidy)` observations across the three trash wheels. Some key variables would be 'homes powered', 'weight' and 'volume' of trash. For available data, total weight of trash collected by professor trash wheel is `r prof_weight` and the total number of cigarette butts collected by Gwynnda in June of 2022 is `r gwynnda_cigs`.

### Problem 3



  










