---
title: "Homework 2"
output: github_document
date: "2026-10-06"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

### Problem 1

**Step 1: Import and Clean Dataset**
```{r read_csv}
##Read and clean the data (copied from assignment)

nytransit_df = read_csv(file = "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                                na = c("", "NA", ".")) #cleans missing data

names(nytransit_df)#overview of all variables

nytransit_df = janitor::clean_names(nytransit_df)#convert to snake case
names(nytransit_df)#checks if above line ran properly
```
Double checking to see if cleaning data done properly:
```{r checking_data2}
nytransit_df
```

**Step 2: Data Manipulation**
```{r select}
##retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character to a logical variable 

nytransit_df <- read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", na = c("", "NA", ".")) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude, 
    route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO"  ~ FALSE
    )
  )
```
Double checking to see if data manipulation done properly:
```{r checking_data}
nytransit_df #entry successfully converted to logical variable, all intended columns kept

nrow(nytransit_df)
ncol(nytransit_df)
```
**About this Dataset:**
(Q: Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?)

This data set (nytransit_df) has the following variables:line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. We first read and tidied the data and then manipulated it so only the desired columns as mentioned were kept as well as converting the 'entry' variables from characters to logical variables. The dataset was cleaned in three steps, first is converting any missing data to 'NA', then using the janitor function to standardize column numbes into snakecase and finally checking the resulting tibble to see if rows need to be skipped to find there were none. The dimension as mentioned in the last checking step is 1868 by 19 (1868 rows or entries by 19 columns or variables). However, at this point, the data is not tidy. This is because the route information is spread across 11 separate columns while in a tidy dataset, each variable forms a column, each obs. has a row and each value has a cell. The route number and route name is still split across 11 columns rather than being tidied into two columns accordingly with their status as two variables and not eleven. 

**Step 3: info retrieval**
```{r retrieval}
#counting distinct stations in nytransit_df
distinct_stations <- nytransit_df |> 
  distinct(station_name, line) |> 
  nrow()

distinct_stations

ada_stations <- nytransit_df |> 
  filter(ada == TRUE) |>   
  distinct(station_name, line) |> 
  nrow()

ada_stations

prop_entrance <- nytransit_df |> 
  filter(vending == "NO") |> 
  summarize(mean(entry))

prop_entrance
```
**Answer the following questions with step 3:**
How many distinct stations are there? 
There are `r distinct_stations` distinct stations.

How many stations are ADA compliant?
`r ada_stations`stations are ADA compliant.

What proportion of station entrances / exits without vending allow entrance?
About `r prop_entrance` of station entrances/exits without vending allow entrance. 

**Step 4: data reformatting**
```{r reformat}
# 1. Reformat data: convert all route columns to character first, then pivot
nytransit_tidy <- nytransit_df |> 
  mutate(across(route1:route11, as.character)) |> 
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  )

# 2. How many distinct stations serve the A train?
a_train_stations <- nytransit_tidy |> 
  filter(route_name == "A") |> 
  distinct(station_name, line) |> 
  nrow()

a_train_stations

# 3. Of the stations that serve the A train, how many are ADA compliant?
a_train_ada_stations <- nytransit_tidy |> 
  filter(route_name == "A", ada == TRUE) |> 
  distinct(station_name, line) |> 
  nrow()

print(a_train_ada_stations)

nytransit_tidy#this checks that the dataset has truly been tidied and it has been because route_number and route_name have been sorted into two columns that corresponds to those two variables rather than across the 11 columns previously

```
**Reformatted data interpretation:**
60 distinct stations serve the A train and of stations that serve the A train, 17 are ADA compliant. 


### Problem 2

**Step 1: Read and clean Mr. Trash Wheel Sheet:**
```{r import}
#Read and clean the Mr. Trash Wheel sheet:specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel, use reasonable variable names, omit rows that do not include dumpster-specific data, round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer) (copied from assignment)

mrtw_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                     sheet = "Mr. Trash Wheel",
                     range = "A2:N742",
                     na = c("", "NA", "."))|>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(sports_balls = as.integer(round(sports_balls)),
         trash_wheel = "Mr. Trash Wheel",
         year = as.numeric(year),
         plastic_bottles = as.numeric(plastic_bottles)
         )

mrtw_df #this is to check for non-data entries, we see a '...15' and '...16' columns which should be ommitted, we also see that the last row doesn't include dumpster-specific data (These 'issues' are now fixed, we also have a clean dataset)
```
**Step 2: Read and clean Prof. Trash Wheeland Gwynnda Sheet:**
```{r import2}
#Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda (copied from assignment)

proftw_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                     sheet = "Professor Trash Wheel",
                     range = "A2:M139",
                     na = c("", "NA", "."))|>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Professor Trash Wheel",
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles)
  )

proftw_df #this is to check cleanliness of df,prof tw has no sports_balls column as a note

gwynnda_df = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
                     sheet = "Gwynnda the Good Wheel of the W",
                     range = "A2:L394",
                     na = c("", "NA", "."))|>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    trash_wheel = "Gwynnda",
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles)
  )

gwynnda_df #this is to check cleanliness of df,gwynnda has no sports_balls column as a note
```

**Step 3: Combine the three datasets**
```{r combine}
#To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining. This new variable is trash_wheel. 
tw_tidy <- bind_rows(mrtw_df, proftw_df, gwynnda_df) |>
  relocate(trash_wheel)

#View combined dataset
tw_tidy #shows a tidied, combined dataset with the variable #trash_wheel showing which dumpster the data belonged to
```

**Step 4: Info retrieval**
```{r retrieval_p2}
# 1. Total weight of trash collected by Professor Trash Wheel (in tons)
prof_weight <- tw_tidy |> 
  filter(trash_wheel == "Professor Trash Wheel") |> 
  summarize(total_weight_tons = sum(weight_tons, na.rm = TRUE))

prof_weight

# 2. Total cigarette butts collected by Gwynnda in June 2022
gwynnda_cigs <- tw_tidy |> 
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"  
  ) |> 
  summarize(total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE))

gwynnda_cigs
```
## About the data: 
This new dataset is `r nrow(tw_tidy)`rows by `r ncol(tw_tidy)` columns and contains `r nrow(tw_tidy)` observations across the three trash wheels. Some key variables would be 'homes powered', 'weight' and 'volume' of trash. For available data, total weight of trash collected by professor trash wheel is `r prof_weight` and the total number of cigarette butts collected by Gwynnda in June of 2022 is `r gwynnda_cigs`.

### Problem 3
**Step 1: Import and Clean Dataset**
```{r read_csvmci}
##Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). (copied from assignment)

mcibaseline_df = read_csv(file = "data/MCI_baseline.csv",
                                skip = 1,
                                na = c("", "NA", "."))#cleans missing data

mcibaseline_df = janitor::clean_names(mcibaseline_df)#convert to snake case

mcibaseline_tidy <- mcibaseline_df |> 
  # 1. Remove participants who already had MCI at baseline
  filter(is.na(age_at_onset) | age_at_onset > current_age) |> 
  # 2. Convert sex and apoe4 numeric codes into categorical factors
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    apoe4 = factor(apoe4, levels = c(0, 1), labels = c("Non-carrier", "Carrier"))
  )

mcibaseline_tidy#checks if above lines ran properly
```
## Discussion: 
There were several key steps when importing this dataset. First is realizing that the first row should be skipped because it doesn't contain column names but variable key instead. After which you go through the tidying steps of converting missing data to 'na', and standardizing variable names into snakecase. Then you go on to filter those who do not have MCI at  baseline before converting sex and apoe4 from 'double' into 'categoricals'. We could probably do logical but the standard is categoricals and it's easier to read/interpret as well. `r mci_developers <- mcibaseline_tidy |> filter(!is.na(age_at_onset)) |> nrow()` were recruited while `r nrow(mcibaseline_tidy)` develop MCI. The average baseline age is `r mcibaseline_tidy |> mean(current_age)` while the proportion of women in the study who are APOE4 carriers is `r mcibaseline_tidy |> filter(sex == "Female") |> summarize(p = mean(apoe4 == "Carrier")) |> pull(p) |> round(3)`.

**Step 2: Import and Clean Dataset 2 electric boogaloo**
```{r read_csvmci2}
##Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

mciamy_df = read_csv(file = "data/mci_amyloid.csv",
                                skip = 1,
                                na = c("", "NA", "."))#cleans missing data

mciamy_tidy = janitor::clean_names(mciamy_df)#convert to snake case

mciamy_tidy# Shows first row skipped so that the actual column names can become first row and also converted everything to snakecase
```
**Step 3: Parse and combine the datasets!**
```{r combine3}
#1. Check who's only in baseline and who's only in amyloid datasets
only_baseline <- anti_join(mcibaseline_tidy, mciamy_tidy, by = c("id" = "study_id"))
only_amyloid  <- anti_join(mciamy_tidy, mcibaseline_tidy, by = c("study_id" = "id"))

#Print counts of non-matching people
nrow(only_baseline)
nrow(only_amyloid)

#2. Combine the datasets (only participants in both kept)
mci_combined <- inner_join(
  mcibaseline_tidy, 
  mciamy_tidy, 
  by = c("id" = "study_id")
)

mci_combined #this is to give the combined dataset a good lookover

#3. Export result to csv!
write_csv(mci_combined, "data/mci_combined.csv")
```
## Final Discussion: 

The longitudinal amyloid dataset contains repeated measurements of the amyloid β 42/40 ratio across follow-up time points (baseline, time_2, time_4, time_6, and time_8).
Using the antijoin function, we could parse out participants who only appeared in either the baseline dataset `r nrow(only_baseline)` or the amyloid dataset `r nrow(only_amyloid)`. Since such participants existed, the inner join function was used to join the two datasets so that only the rows that have matching values according to the study_id were kept. The resulting combined tibble contains `r ncol(mci_combined)` columns and `r nrow(mci_combined)` rows. The columns start with the baseline dataset's variables on the left before going into the amyloid dataset without repeating the study_id column which would be redundant. The number of retained participants is `r nrow(mci_combined)`. Compared to the `r nrow(mcibaseline_tidy)` observations from baseline and `r nrow(mciamy_tidy)`, we see that `r nrow(mcibaseline_tidy) - nrow(mci_combined)` and `r nrow(mciamy_tidy) - nrow(mci_combined)` have been deleted for the combination, respectively. 


  










